---
title: "Auditing a quality filter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditing a quality filter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4
)
set.seed(7)
```

```{r setup}
library(TrustworthySDM)
```

## The question

You have occurrence records. Some have good coordinates and some do not. You are
about to throw the bad ones away, because that is what one does.

The question this package asks is not *which* records to throw away. It is: once
you have thrown them away, **what did you do to the environmental distribution
your model is about to learn from?**

If record quality is unrelated to the environment, the answer is *nothing*, and
you can stop reading. If it is not — and it usually is not, because
georeferencing precision depends on terrain and access and era, all of which vary
across environmental space — then filtering has quietly reweighted environmental
space, and the model will learn a displaced niche.

## A worked example

`sim_occ()` simulates records in river basins. Elevation carries a basin-level
random effect; temperature and precipitation follow elevation; and coordinate
uncertainty is generated to depend on elevation too, so that quality and
environment are coupled through terrain, as they are in real data.

```{r}
d <- sim_occ(n = 1500, seed = 7)
str(d[, c("bio1", "bio12", "elev", "basin", "coord_uncertainty")])
```

It ships four filters, and the whole point of the object is that they should get
*different* verdicts.

```{r}
colMeans(d[, c("keep_random", "keep_precision", "keep_outlier", "keep_strict")])
```

## A filter that does nothing

`keep_random` drops records at random. This is the negative control: an audit
that flags it is firing indiscriminately and is worthless.

```{r}
env <- c("bio1", "bio12", "elev")

audit(d, quality = "keep_random", env = env, spatial_block = "basin")
```

`env-AUC` sits at 0.5. Retention carries no environmental information, so
filtering and reweighting will point the same way, and the filtering decision is
not carrying your result.

One caveat, and it is in the printed advice for a reason: a null here is a
*failure to resolve structure*, not a proof of its absence. With few spatial
blocks it is a cross-validation null and nothing more.

## A filter that does something

`keep_precision` drops records with coarse coordinates — and coordinate
precision tracks elevation. This is the ordinary case: a filter defined on
metadata that projects onto the environment without anyone intending it to.

```{r}
a <- audit(d,
  quality = "keep_precision", env = env,
  metadata = c("year", "source"), spatial_block = "basin"
)

summary(a)
```

Read it in order.

**Which axes moved.** The retained records sit higher, colder and wetter. That
is not a fact about the species; it is a fact about where a coordinate can be
pinned down.

**Did the niche move at all.** The energy distance rejects the null that the
retained records are a random environmental subsample.

**Environmental, or provenance?** This is what `metadata` buys you. A metadata
model recovers retention well, which is unsurprising — the filter *was* a
metadata rule. The number that matters is the incremental one: environmental
structure that survives controlling for who collected the record and when. If it
vanished, the shift would be a provenance artefact. It does not vanish.

**Can it be undone?** The effective sample size of the inverse-propensity
weights is a healthy fraction of the retained records, so reweighting is
supported by the data here.

```{r, fig.alt = "Left: per-axis standardized mean differences. Right: propensity score distributions for retained and discarded records."}
plot(a)
```

## Both ends of the bracket

The verdict is `structured`, and the advice is to fit *both* corrections.

```{r}
w_ipw <- ipw_weights(a) # undoes the selection
w_trust <- trust_weights(a) # soft version of the filter

c(ipw_ess = attr(w_ipw, "ess"), retained = a$n_high)
```

These do not disagree by accident. Filtering downweights the environments where
low-quality records concentrate; inverse-propensity weighting upweights the rare
retained records in those same environments, to undo the selection. They move the
niche in **opposite** directions and bracket the truth between them.

Which end is right depends on whether coordinate error is *directed* — whether
mislocated records are pushed systematically toward particular environments, or
merely scattered — and that is not identifiable from occurrence data. Fitting
with one set of weights and reporting the result commits your analysis to one end
of a bracket you never measured.

Fit both. Report the spread.

## The filter that hides

`keep_outlier` is a Mahalanobis-distance screen on the environment: it removes
records that sit far out in environmental space. Screens of this kind are
standard, and they look conservative.

They are the opposite of conservative, and they are almost invisible.

```{r}
o <- audit(d, quality = "keep_outlier", env = env, spatial_block = "basin")

o$shift[, c("variable", "smd", "log_vr")]
```

Every standardized mean difference is near zero. The filter is *symmetric*: it
trims both tails, so it moves no mean. An audit built on mean shifts would report
this filter as benign, and an energy test may well fail to reach significance on
it.

Look at the variance ratio instead. The retained records have collapsed
variance on every axis. The filter has not moved the niche — it has **shrunk**
it, and a model fitted on the survivors will underestimate the environmental
breadth of the species.

The propensity model catches it, but only because it carries quadratic terms:

```{r}
lin <- audit(d,
  quality = "keep_outlier", env = env, spatial_block = "basin",
  control = audit_control(quadratic = FALSE)
)

c(linear = lin$propensity$auc_env, quadratic = o$propensity$auc_env)
```

The linear model is not merely uninformative here. Out of fold it is
*anti-predictive*: it discriminates in the wrong direction, because the only
linear signal it can find is fold-specific noise. That is why the shift screen in
this package tests `|AUC − 0.5|` rather than `AUC > 0.6`. A one-sided test would
have looked at an AUC of 0.42, concluded "no structure", and told you your
niche-deflating filter was benign.

Keep `quadratic = TRUE`.

## The filter that cannot be undone

`keep_strict` is a near-deterministic certification rule: records below a terrain
threshold are almost never certified. A handful are anyway.

```{r}
s <- audit(d, quality = "keep_strict", env = env, spatial_block = "basin")

s$flag
c(
  ess_trimmed = s$positivity$ess_ratio,
  ess_untrimmed = s$positivity$ess_ratio_untrimmed
)
```

The shift is real and large. But the correction is carried by a few records with
enormous weights, because there are environments the filter *emptied*, and no
weight can conjure records back into an empty environment. This is a positivity
failure, and it is the limit of what reweighting can do.

Notice how far apart the trimmed and untrimmed effective sample sizes are.
Trimming the weights is what stops those few records from carrying the whole
correction — which means "the effective sample size" is not a well-defined number
until you say whether the weights were trimmed first. Both are reported here for
that reason. If you quote one, say which.

Here, the right move is not to weight and move on. It is to report the
filter-to-IPW bracket, and to say plainly that the data cannot resolve which end
of it is correct.

## Bringing your own data

```{r, eval = FALSE}
# from a continuous coordinate-uncertainty field, as GBIF supplies it
occ$high <- as_quality(occ$coordinateUncertaintyInMeters, threshold = 100)

# or from CoordinateCleaner's .summary column
occ$high <- as_quality(cc_output)

audit(occ,
  quality = "high",
  env = c("bio1", "bio12", "bio15"),
  metadata = c("year", "institutionCode"),
  spatial_block = "basin"
)
```

When quality is continuous there is no single correct cut, and auditing at one
threshold answers a question you chose rather than a question about your data.
Sweep it. If the shift *grows* as the cut gets stricter, then cleaning harder is
making the bias worse, and there is no bias-free threshold to find.
